Synthesizing Symmetric Lenses

### 2018-07-10 papers/SymmetricLenses/main.tex

A natural way to expand the scope of such tools is to consider a more
exrpressive language of lenses.  For example, \emph{symmetric
lenses}~\cite{symmetric-lenses} encompass a broad collection of bidirectional
transformations that include bijections as well as more flexibile transformations that
allow each side to retain information not present
in the other.  If $\ell$ is a symmetric lens between formats $S$ and
$T$, then $\ell$  can transform data from $S$ to $T$ or vice versa
equally well.  A user would typically 
run $\ell$ in the forwards direction (from $S$ to $T$) when a user updates
a string $s \in S$ to propagate those updates to $T$.
However, because data in format $S$ may contain some information not present in format $T$,
the $S \rightarrow T$ transformation takes
\emph{both} the edited string $s \in S$ and a string $t \in T$ as arguments.
Inuitively, the lens extracts the data ``shared'' between the two
formats from $s$
and uses $t$ to supply any missing information.
% required by $T$ that is missing
%as it generates an output $t'$.  If $s$ is missing information, the transformation should
%extract the missing components from $t$.

@@

The second difficulty with synthesizing symmetric lenses is more
fundamental: While the number of bijective lenses between two
related formats is often tiny, the number of symmetric lenses is often large. In
practice, the difference in cardinality means that when a search algorithm
selects the first bijective lens it finds, that lens has a good chance of being
the one a user wants. In contrast, and to our surprise, our experiments show
that if our search algorithm selects the first symmetric lens it finds, it
typically fails to find the lens the user wants. The flexibity these lenses
admit when it comes to eliding information generates an explosion of possible
transformations. To disambiguate between the many possible lenses, we need new
principles for identifying the ``likely'' ones, and we need new algorithms to
search the space of program transformations using these principles.

To overcome this second difficulty, we turned to information theory to try to
find a way to characterize the likely and unlikely lenses. Intuitively, a
``likely'' lens is one that propagates a lot of information from source to
target and also from target back to source. Conversely, given the source, the
target is ``unlikely'' to require very much additional information in order to
reconstruct it, under the assumption that users typically want to synchronize
formats with a large amount of shared information. In order to implement this
intuition, we turn to \emph{stochastic regular expressions} (SREs). Stochastic
regular expressions define a set of strings, as do ordinary regular expressions,
but they also provide the expected distribution of strings within that set. Such
distributions allow us to compute the amount of information gained when
presented an element of the set. For example, the language of the SRE
$\lstinline{"a"} |_{0.5}
\lstinline{"b"}$ contains two elements (the strings \lstinline{"a"} and \lstinline{"b"}).
The probability 0.5 annotating the union indicates that both elements are
equally likely to appear. Consequently, if we observe an element of that
language (either \lstinline{"a"} or
\lstinline{"b"}), we will gain one bit of information. Said another way, one bit, together with
the SRE, suffices to recover the string in question. Interestingly, if the
probability annotation were anything but 0.5, the expected information content
(or entropy) would actually be less than 1 (given a stream of such choices, on
average we could encode the the contents of stream in fewer bits than the the
number of choices present). On the other hand, consider the language of the SRE
\lstinline{c}. It contains just one element (the string $c$). If we observe that
our algorithm generates $\lstinline{"c"}$, we gain zero bits of information ---
the string was already determined. Using this foundation, a lens $\Lens$ that
synchronizes data between $\Regex$ and $\RegexAlt$ is ``likely,''
in a technical sense, if given a string $s$ in the source $S$, very little information
is needed to determine which synchronized string $t$ is present in the target
$T$.

@@

>>>>

A natural way to expand the scope of such tools is to consider a more
expressive language of lenses.  For example, \emph{symmetric
lenses}~\cite{symmetric-lenses} encompass a broad collection of bidirectional
transformations that include bijections as well as more flexibile transformations that
allow each side to retain information not present
in the other.  If $\ell$ is a symmetric lens between formats $S$ and
$T$, then $\ell$  can transform data from $S$ to $T$ or vice versa
equally well.  A user would typically 
run $\ell$ in the forwards direction (from $S$ to $T$) when a user updates
a string $s \in S$ to propagate those updates to $T$.
However, because data in format $S$ may contain some information not present in format $T$,
the $S \rightarrow T$ transformation takes
\emph{both} the edited string $s \in S$ and a string $t \in T$ as arguments.
Inuitively, the lens extracts the data ``shared'' between the two
formats from $s$
and uses $t$ to supply any missing information.
% required by $T$ that is missing
%as it generates an output $t'$.  If $s$ is missing information, the transformation should
%extract the missing components from $t$.

@@

The second difficulty with synthesizing symmetric lenses is more
fundamental: While the number of bijective lenses between two
related formats is often tiny, the number of simple symmetric lenses is much larger. In
practice, the difference in cardinality means that when a search algorithm
returns the first bijective lens it finds, that lens has a good chance of being
the desired lens. In contrast, and to our initial surprise, experiments showed
that if a naive search algorithm selects the first simple symmetric lens it
finds, it is typically \emph{not} the desired lens.  The flexibility
that such lenses provide in eliding information 
generates an explosion of possible transformations. To disambiguate
between the many possible lenses, we need new 
principles for identifying the ``likely'' ones, and we need new
algorithms guided by those principles to
search the space of program transformations.

To overcome this second difficulty, we turn to information theory 
to characterize the likely and unlikely lenses. 
We consider a ``likely'' lens to be one that propagates a lot of
information from source to 
target and vice versa. Conversely, an ``unlikely'' lens requires a
large amount of additional information to recover the target given a
source and vice versa.  
This categorization follows from the intuition
that users will typically want to synchronise formats with lots
of shared information, and so lenses that preserve 
more information are preferable.

To implement this
intuition, we turn to \emph{stochastic regular expressions} (SREs). Stochastic
regular expressions define a set of strings, as do ordinary regular expressions,
but they also provide the expected distribution of strings within that set. Such
distributions allow us to compute the amount of information gained when
presented an element of the set.

@@

\ksf{Consider integrating more detailed discussion of SREs that was
formerly in the intro.  Commented out here for now.}
%%To implement this intuition, we turn to \emph{stochastic regular
%%expressions} (SREs). Stochastic 
%%regular expressions define a set of strings, as do ordinary regular expressions,
%%but they also provide the expected distribution of strings within that set. Such
%%distributions allow us to compute the amount of information gained when
%%presented an element of the set. For example, the language of the SRE
%%$\lstinline{"a"} |_{0.5}
%%\lstinline{"b"}$ contains two elements (the strings \lstinline{"a"} and \lstinline{"b"}).
%%The probability 0.5 annotating the union indicates that both elements are
%%equally likely to appear. Consequently, if we observe an element of that
%%language (either \lstinline{"a"} or
%%\lstinline{"b"}), we will gain one bit of information. Said another way, one bit, together with
%%the SRE, suffices to recover the string in question. Interestingly, if the
%%probability annotation were anything but 0.5, the expected information content
%%(or entropy) would actually be less than 1 (given a stream of such choices, on
%%average we could encode the the contents of stream in fewer bits than the the
%%number of choices present). On the other hand, consider the language of the SRE
%%\lstinline{c}. It contains just one element (the string $c$). If we observe that
%%our algorithm generates $\lstinline{"c"}$, we gain zero bits of information ---
%%the string was already determined. Using this foundation, a lens $\Lens$ that
%%synchronizes data between $\Regex$ and $\RegexAlt$ is ``likely,''
%%in a technical sense, if given a string $s$ in the source $S$, very little information
%%is needed to determine which synchronized string $t$ is present in the target
%%$T$.
